Epoch of Reionization Power Spectrum Results from PAPER-128

### 2016-08-05 PAPER.tex
fleshed out some OQE stuff. Started adding cites

Galaxy observations have also begun to place constraints on the EoR. With their high luminosities, quasars make for powerful probes of the state of the IGM through absorption line studies. As photons from a quasar are redshifted, they may be absorbed by neutral hydrogen at the Lyman-$\alpha$ line, producing Gunn-Peterson trough features in quasar spectra. Observations of these spectra in high redshift galaxies have signalled the end of reionization by $z ~ 6$, and experiments are continuously pushing galaxy observations to higher redshifts.

@@

Major existing 21\,cm experiments include both those targeting to measure the sky-averaged `global' 21\,cm signal and those employing interferometric elements and the use of cross-correlations and statistical power-spectral measurements. Examples of the former are EDGES, the LWA, LEDA, DARE, BIGHORNS, and SARAS. Major interferometers include the GMRT, LOFAR, the MWA, the 21CMA, and PAPER.

In this paper we present results from the $128$ antenna configuration of the Donald C. Baker Precision Array for Probing the Epoch of Reionization (PAPER). PAPER is a highly redundant array consisting of dual-polarization dipole antennas. Its configuration is optimal for boosting power spectrum sensitivity and the analysis used relies on a foreground avoidance approach where we measure power spectrum values in a region of delay space that is uncontaminated by foregrounds. Previous upper limits on the EoR power spectrum include those made by PAPER-32, which used its redundant configuration to its advantage to obtain an upper limit of $\Delta^{2}(k) \leq (41 {\rm mK})^{2}$ at $k=0.28h {\rm Mpc}^{-2}$. Doubling in array size and employing an updated analysis pipeline that included fringe-rate filtering and an optimal quadratic estimator framework for the power spectrum analysis achieved an upper limit on $\Delta^{2}(k)$ of $(22.4 {\rm mK}))^{2}$ in the range $0.15 < k < 0.5h {\rm Mpc}^{-1}$ at $z = 8.4$. The result from PAPER-64 was the best published upper limit thus far, with PAPER-128 achieving a factor of ?? in improvement.

@@

The PAPER signal chain consists of drift-scan antenna elements...
The PAPER beam is...

@@

For our analysis, we use the package {\sc Omnical} for redundant calibration. Similarly to the analysis of PAPER-64, {\sc Omnical} comprises of three steps. The first is {\tt FirstCal}, which uses all baseline redundancies to generate a static gain solution for each antenna that will unwrap any phase wrapping between two identical baselines. We perform {\tt FirstCal} because the next stage of {\sc Omnical} cannot tell the difference between a phase of $0$ and $2\pi$, for example. The second step is {\tt LogCal}, which takes the log of all the visibility equations and separates the real and imaginary components into two matrices. Coarse solutions are determined for both the antenna gains and `model' visibilities (one for each baseline type) simultaneously. The final step of {\sc Omnical} is {\tt LinCal},, which applies small perturbations to the {\tt LogCal}, solutions in an iterative fashion, honing in on the optimal solutions.

@@

\subsection{Delay-Domain Foreground Removal}

@@

In practice, we filter our calibrated visibilities by de-convolving out our sampling function (which contains flags due to RFI) from our delay-domain visibilities using a CLEAN-like algorithm and restricting our clean components to inside the horizon limit (set by the baseline length) plus a $15 {\rm ns}$ buffer. The Fourier-transformed clean components are then subtracted from our visibilities. This filtering process is performed on a per-baseline, per-integration basis, and we achieve a brightness suppression of $~2$ orders of magnitude in our visibilities.

@@

We perform a final round of RFI-removal by flagging visibilities that lie more than $4\sigma$ above the mean on a time, frequency, and baseline basis. \cc{we flag specific channels before delay filtering too, to avoid leakage outside of the horizon limit due to Fourier-transforming super un-smooth signals}

Finally, we stack our data in LST into two datasets, alternating between even and odd Julian Dates to create an `even' and `odd' LST-binned dataset. \cc{other jack-knife tests...}

@@

It was shown in \citet{parsons_et_al2016} that filtering in fringe-rate space is equivalent to weighting data by the primary beam. Because of this, we determine an optimal fringe-rate filter that up-weights parts of the sky that our beam is most sensitive to, while down-weighting the opposite. We also filter out the zero fringe-rate bin, which contains signals that are constant in time, such as crosstalk. We then apply the filter on a per-baseline basis as a convolution in the time-domain, which has the affect of averaging our data in time.

Using an optimal fringe-rate filter comes at the cost of signal loss due to a reduction of the number of independent time samples in our dataset...

@@

With our final visibilities in hand, we are ready to estimate the 21\,cm power spectrum, $\hat{\textbf{p}}(\textbf{k})$. The power spectrum measurement is a function of $\textbf{k} = (k_{x},k_{y},k_{z})$, where each component of $k$ represents the Fourier-dual to spatial coordinates $(x,y,z)$. Our visibilities have already been Fourier-transformed along two of these axes, namely, along $k_{\perp} = \sqrt{k_{x}^{2}+k_{y}^{2}}$. This perpendicular component of $k$ is analogous to the spatial Fourier-modes in the plane of the sky, while the parallel component of $k$, $k_{\parallel}$, is the Fourier-transform of our visibilities along the line-of-sight (in the frequency direction).

@@

\noindent where $\tilde{T}_{b}(\textbf{k})$ is the spatial Fourier-transform of the brightness temperature of the sky, $\delta^{D}$ is the Dirac delta function, and $\langle \rangle$ is an ensemble average. As a simplistic explanation, we need to take our visibility temperature measurements, Fourier-transform them along the frequency axis and convert from units of Jys to Kelvin, square, and bin in constant contours of $k$. In practice, however, we use linear algebra techniques as explained in the following section in order to obtain both $\hat{\textbf{p}}(\textbf{k})$ and its associated error bars.

@@

\noindent represents power spectrum quantities (un-normalized) for each $k$ bin indexed by $\alpha$. Here we see two copies of our data vector \textbf{x} and two copies of $\textbf{C}^{-1}$ where $\bf{C} \equiv \langle\textbf{x}\textbf{x}^{\dagger}\rangle$, the true covariance matrix of \textbf{x}. The operator $\textbf{Q}_{\alpha}$ transforms our visibilities into power spectrum space by performing a Fourier transform along the frequency axis, and $b_{\alpha}$ is a noise bias that is subtracted off.

In practice, $\bf{x}$ has dimensions of time versus frequency and exists for every baseline (in a given redundant group), and group (such as an `even' or `odd' LST-binned set). Notice that there is a copy of $\textbf{C}^{-1}\textbf{x}$ on each side of the expression for $\hat{q}_{\alpha}$. When performing this computation, we do not allow the cross-multiplication to occur if both quantities are derived for the same baseline, or the same group. For example, we do not allow $(\textbf{C}^{-1}\textbf{x})_{even'A}$ to be multiplied with $(\textbf{C}^{-1}\textbf{x})_{even,B}$ since both quantities are formed from the `even' dataset, nor do we allow $(\textbf{C}^{-1}\textbf{x})_{even,A}$ to be multiplied with $(\textbf{C}^{-1}\textbf{x})_{odd,A}$, since both quantities are formed for baseline $A$. We do not include these cases in order to eliminate any biases from allowing auto power from identical baselines or groups to remain.

@@

\noindent The window function is constructed so that each row sums to unity. This ensures proper normalization of the power spectrum estimates. In practice, we have a choice of \textbf{M} and in our analysis choose to use a method where it is determined from the Cholesky decomposition of \textbf{F}... \cc{this may change...}.

@@

\noindent and there is no signal loss associated with this computation. The expression for the output power spectrum, $P_{out}$, is comprised of two terms - the covariance treated power spectrum associated with \textbf{r}, and that of \textbf{x} alone.

@@

>>>>

Galaxy observations have also begun to place constraints on the EoR. With their high luminosities, quasars make for powerful probes of the state of the IGM through absorption line studies. As photons from a quasar are redshifted, they may be absorbed by neutral hydrogen at the Lyman-$\alpha$ line, producing Gunn-Peterson trough features in quasar spectra. Observations of these spectra in high redshift galaxies have signalled the end of reionization by $z \sim 6$, and experiments are continuously pushing galaxy observations to higher redshifts.

@@

Major existing 21\,cm experiments include both those targeting to measure the sky-averaged `global' 21\,cm signal and those employing interferometric elements and the use of cross-correlations and statistical power-spectral measurements. Examples of the former are EDGES (\citealt{bowman2010}), the LWA (\citealt{ellingson_et_al2009}), LEDA (\citealt{greenhill_bernardi2012}), DARE (\citealt{burns2012}), SciHi (\citealt{voytek2014}), BIGHORNS (\citealt{sokolowski2015}), and SARAS (\citealt{patra2015}). Major interferometers include the GMRT, LOFAR, the MWA, the 21CMA, and PAPER.

In this paper we present results from the $128$ antenna configuration of the Donald C. Baker Precision Array for Probing the Epoch of Reionization (PAPER). PAPER is a highly redundant array consisting of dual-polarization dipole antennas. Its configuration is optimal for boosting power spectrum sensitivity and the analysis used relies on a foreground avoidance approach where we measure power spectrum values in a region of delay space that is uncontaminated by foregrounds. Previous upper limits on the EoR power spectrum include those made by PAPER-32, which used its redundant configuration to its advantage to obtain an upper limit of $\Delta^{2}(k) \leq (41 {\rm mK})^{2}$ at $k=0.28h {\rm Mpc}^{-2}$. Doubling in array size and employing an updated analysis pipeline that included fringe-rate filtering and an optimal quadratic estimator framework for the power spectrum analysis achieved an upper limit on $\Delta^{2}(k)$ of $(22.4 {\rm mK}))^{2}$ in the range $0.15 < k < 0.5h {\rm Mpc}^{-1}$ at $z = 8.4$. This result from PAPER-64 was the best published upper limit thus far, with PAPER-128 achieving a factor of ?? in improvement.

@@

The PAPER signal chain consists of drift-scan antenna elements... \cc{mention Casper and instrumentation things}
The PAPER beam is...

@@

For our analysis, we use the package {\sc Omnical} for redundant calibration. Similarly to the analysis of PAPER-64, {\sc Omnical} comprises of three steps. The first is {\tt FirstCal}, which uses all baseline redundancies to generate a static gain solution for each antenna that will unwrap any phase wrapping between two identical baselines. We perform {\tt FirstCal} because the next stage of {\sc Omnical} cannot tell the difference between a phase of $0$ and $2\pi$, for example. The second step is {\tt LogCal}, which takes the log of all the visibility equations and separates the real and imaginary components into two matrices. Coarse solutions are determined for both the antenna gains and `model' visibilities (one for each baseline type) simultaneously. The final step of {\sc Omnical} is {\tt LinCal}, which applies small perturbations to the {\tt LogCal} solutions in an iterative fashion, honing in on the optimal solutions.

@@

\subsection{Delay-Domain Foreground Removal}
\label{sec:fgclean}

@@

In practice, we first flag specific frequency channels that contain known RFI. If this is not done prior to foreground filtering, the EoR window may become contaminated by leakage from the Fourier transform of sharp RFI features. In addition to flagging entire channels with obvious RFI (such as channels occupied by ORBCOMM and the International Space Station) we also flag some with lower levels of RFI caused by radio and other Earth-origin activity.

Next, we filter our calibrated visibilities by de-convolving out our sampling function (which contains flags due to RFI) from our delay-domain visibilities using a CLEAN-like algorithm and restricting our clean components to inside the horizon limit (set by the baseline length) plus a $15 {\rm ns}$ buffer. The Fourier-transformed clean components are then subtracted from our visibilities. This filtering process is performed on a per-baseline, per-integration basis, and we achieve a brightness suppression of $~2$ orders of magnitude in our visibilities.

@@

We perform a final round of RFI-removal by flagging visibilities that lie more than $4\sigma$ above the mean on a time, frequency, and baseline basis. Finally, we stack our data in LST into two datasets, alternating between even and odd Julian Dates to create an `even' and `odd' LST-binned dataset.

@@

\begin{figure*}[!]
	\centering
	%\includegraphics[width=1.0\textwidth]{raw.png}
	\caption{Top: The optimal fringe-rate profile (weighting in fringe-rate space) for our I polarization beam for an E/W baseline separation of $30 m$. Bottom: The time-domain convolution kernel corresponding to the curve in the top panel.}
	\label{fig:frf}
\end{figure*}

It was shown in \citet{parsons_et_al2016} that filtering in fringe-rate space is equivalent to weighting data by the primary beam. Because of this, we determine an optimal fringe-rate filter (Figure \ref{fig:frf}) that up-weights parts of the sky that our beam is most sensitive to, while down-weighting the opposite. We also filter out the zero fringe-rate bin, which contains signals that are constant in time, such as crosstalk. We then apply the filter on a per-baseline basis as a convolution in the time-domain, which has the affect of averaging our data in time.

Using an optimal fringe-rate filter comes at the cost of signal loss due to a reduction of the number of independent time samples in our dataset. This is described in more detail in section \ref{sec:Sigloss}.

@@

With our final visibilities in hand, we are ready to estimate the 21\,cm power spectrum, $\hat{\textbf{p}}(\textbf{k})$. The power spectrum measurement is a function of $\textbf{k} = (k_{x},k_{y},k_{z})$, where each component of $k$ represents the Fourier-dual to spatial coordinates $(x,y,z)$. Our visibilities have already been Fourier-transformed along two of these axes, namely, along $k_{\perp} = \sqrt{k_{x}^{2}+k_{y}^{2}}$. This perpendicular component of $k$ is analogous to the spatial Fourier-modes in the plane of the sky, while the parallel component of $k$, $k_{\parallel}$, is the Fourier-transform of our visibilities along the frequency axis as sampled by our interferometer. It is important to distinguish $k_{\parallel}$, which represents delay-modes, from $\eta$, which is the perfect Fourier transform along frequency. Because of the chromaticity of our instrument, taking the Fourier transform of our data along frequency produces $k_{\parallel}$ modes, and there exists a wedge in $k$-space which is contaminated by foregrounds. In other words, $k_{perp} \propto \textbf{b}/{\lambda}$, and as our baseline length increases, the horizon limit in delay space as described in section \ref{sec:fgclean} also increases, creating a foreground dominated wedge.

@@

\noindent where $\tilde{T}_{b}(\textbf{k})$ is the spatial Fourier-transform of the brightness temperature of the sky, $\delta^{D}$ is the Dirac delta function, and $\langle \rangle$ is an ensemble average. As a simplistic explanation, we need to take our visibility temperature measurements, Fourier-transform them along the frequency axis and convert from units of Jys to Kelvin, square, and bin in constant contours of $k$. This can be represented mathematically by

\begin{equation}
\hat{\textbf{p}} = \Big(\frac{\lambda^{2}}{2k_{B}}\Big) \frac{X^{2}Y}{\Omega^{pp}B} | \tilde{V} (\textbf{u},\eta) |^{2}
\end{equation}

\noindent where $k_{B}$ is the Boltzmann constant, $X$ and $Y$ are factors that relate the cosmological Fourier coordinates to interferometric coordinates by $(Xk_{x},Xk_{y}, Yk_{z}) \equiv 2\pi(u,v,\eta)$, $\Omega_{pp}$ is the integral of the primary beam squared over frequency and angle, and $\tilde{V}$ is the Fourier transformed visibilities.

In practice, however, we use linear algebra techniques as explained in the following section in order to obtain both $\hat{\textbf{p}}(\textbf{k})$ and its associated error bars.

@@

\noindent represents power spectrum quantities (un-normalized) for each $k$ bin indexed by $\alpha$. Here we see two copies of our data vector \textbf{x} and two copies of $\textbf{C}^{-1}$ where

\begin{equation}
\textbf{C} \equiv \langle\textbf{x}\textbf{x}^{\dagger}\rangle,
\end{equation}

\noindent the true covariance matrix of \textbf{x}. The operator $\textbf{Q}_{\alpha}$ transforms our visibilities into power spectrum space by performing a Fourier transform along the frequency axis, and $b_{\alpha}$ is a noise bias that is subtracted off.

Our data $\bf{x}$ has dimensions of time versus frequency and exists for every baseline (in a given redundant group), and group (such as an `even' or `odd' LST-binned set). Notice that there is a copy of $\textbf{C}^{-1}\textbf{x}$ on each side of the expression for $\hat{q}_{\alpha}$. When performing this computation, we do not allow the cross-multiplication to occur if both quantities are derived for the same baseline, or the same group. For example, we do not allow $(\textbf{C}^{-1}\textbf{x})_{even,A}$ to be multiplied with $(\textbf{C}^{-1}\textbf{x})_{even,B}$ since both quantities are formed from the `even' dataset, nor do we allow $(\textbf{C}^{-1}\textbf{x})_{even,A}$ to be multiplied with $(\textbf{C}^{-1}\textbf{x})_{odd,A}$, since both quantities are formed for baseline $A$. We do not include these cases in order to eliminate any biases from allowing auto power from identical baselines or groups to remain.

@@

\noindent The window function is constructed so that each row sums to unity. This ensures proper normalization of the power spectrum estimates. In practice, we have a choice of \textbf{M} as described in \citet{liu_et_al2014b} and in our analysis choose to use a method where it is determined from the Cholesky decomposition of \textbf{F}. Namely, we decompose the Fisher matrix such that $\textbf{F} = \textbf{L}\textbf{L}^{\dagger}$, where $\textbf{L}$ is a lower triangular matrix. Next, we construct $\textbf{M} = \textbf{D}\textbf{L}^{-1}$, where \textbf{D} is a diagonal matrix. Hence, $\textbf{W} = \textbf{D}\textbf{L}^{\dagger}$ and is an upper triangular matrix. This window function was constructed in a way to prevent the leakage of foreground power from low $k$ to high $k$ modes. Specifically, we order the elements in $\textbf{F}$ in such a way so that power can leak from high $k$ modes to low $k$ modes, but not vice versa. Since most foreground power shows up at low $k$'s, this method ensures a window function that retains clean, noise-dominated measurements.

@@

\noindent and there is no signal loss associated with this computation. The expression for the output power spectrum, $P_{out}$, is comprised of two terms: the covariance treated power spectrum associated with \textbf{r}, and that of \textbf{x} alone.

@@

\subsection{Jack-Knifes}
\label{sec:jack}
